---
title: "Literate Analysis with Ocean Water Samples"
subtitle: "coreR Session 05 — Section 5.7 Practice"
format:
  html:
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

# Introduction

This Quarto document follows the **Section 5.7 practice exercise** from coreR *Session 05 (Literate Analysis with Quarto)*: building a simple **prose + code** analysis using real seawater chemistry data (`BGchem2008data.csv`).\
Source: <https://learning.nceas.ucsb.edu/2024-10-coreR/session_05.html>

## About the data

The seawater chemistry dataset used here is `BGchem2008data.csv` (downloaded previously in the course).

-   Date downloaded: August 8, 2025
-   Download source: <https://arcticdata.io/metacat/d1/mn/v2/object/urn%3Auuid%3A35ad7624-b159-4e29-a700-0c0770419941>

You will explore the nitrogen to phosphate ratio, which indicates the health of the ocean ecosystem. <https://beal-agulhas.earth.miami.edu/teaching/courses/lecture-five/index.html>

## My About Section

The redfield ratio is a statistic that describes the comparative levels of the three vital nutrients carbon, nitrogen, and phosphorus in oceanic environments. These three elements are found in roughly a ratio of 106:16:1 (C:N:P), and a measured ratio near this value indicates a healthy ocean ecosystem that has a functioning flow of nutrients through the environment.

Furthermore, the ratio can be simplified to just N:P, with a desired ratio of 16:1. The further from this ratio the more unhealthy the ecosystem is, which can be caused by different factors such as pollution, climate change, or simply changing of ocean currents. In any case, change in the Redfield ratio means that more research in the area is required, since this can have significant effects on the local species as well as the global ecosystem.

## Setup

Best practice: put all `library()` calls in one chunk near the top so collaborators can see what packages are required.

```{r}
library(here)
library(readr)
```

# Read in the data

Quarto runs R in the same way as a script: objects only exist after the chunks that create them have been run.

Using `here()` is a convenient way to build paths relative to your project root (e.g., your `.qmd` might live in `scripts/` while data lives in `data/`).

```{r}
bg_chem <- read_csv(here::here("assignments/water_sample_data/Raw", "BGchem2008data.csv"))
```

(Optional) Quick peek at the data:

```{r}
# Uncomment if you want more inspection output
 names(bg_chem)
 #str(bg_chem)
 head(bg_chem)
 #summary(bg_chem)
```

# Analysis

As the “analysis”, we’ll calculate a few simple summary statistics and examine how the nitrogen:phosphate values compare to the **Redfield ratio** concept.

## Calculate summary statistics

Compute mean nutrient values (using `na.rm = TRUE` so missing values don’t break the calculation).

```{r}
nitrate <- mean(bg_chem$NO3, na.rm = TRUE)
nitrite <- mean(bg_chem$NO2, na.rm = TRUE)
amm     <- mean(bg_chem$NH4, na.rm = TRUE)
phos    <- mean(bg_chem$P,   na.rm = TRUE)
```

## Calculate mean Redfield ratio

```{r}
ratio <- (nitrate + nitrite + amm) / phos
ratio
```

You can report results inline in Quarto like this:

The Redfield ratio for this dataset is approximately: `r round(ratio)`

## Plot Redfield ratio measurements

Finally, create a simple base R plot that shows the relationship between phosphate and the total nitrogen components.

```{r}
plot(
  bg_chem$P,
  bg_chem$NO2 + bg_chem$NO3 + bg_chem$NH4,
  xlab = "Phosphate (P)",
  ylab = "Total nitrogen (NO2 + NO3 + NH4)",
  main = "Ocean water samples: total N vs P"
)
```

## My Graph Analysis

This graph shows a roughly linear trend, where more phosphorus is linearly correlated with nitrogen. This shows that phosphorus concentrations are highly dependent on nitrogen concentrations. Interestingly, most of these points along this trend do not have a healthy Redfield ratio, with nearly all of the observations having a ratio far below 16:1. This indicates that the area these observations were collected in have a nutrient poor environment, and further analysis needs to be performed to find out why this is the case.

# Conclusion

In this mini-analysis, we:

-   Loaded packages in a dedicated setup chunk
-   Read in `BGchem2008data.csv` using a project-root-relative path (`here()`)
-   Calculated simple nutrient means
-   Computed a mean nitrogen:phosphate ratio
-   Created a basic plot relating total nitrogen to phosphate

This is the core idea of **literate analysis**: combine narrative and code in one document so the workflow is easier to understand and reproduce.
