---
title: "a08"
editor: visual
author: "Andreas Krumbein"
---

## Data

The data used refers to the boundaries of Alaska's Salmon and People (SASAP) regions to reflect watershed regions and high resolution visualization of the coastline. Population data across Alaska was obtained from a study identifying population and languages spoken across communities in Alaska. River location data was obtained from a study that adapted data from the United States Geological Survey Hydrographic geodatabase and merged this file with a file referring to the regional boundaries of Alaska to visualize rivers across the state of Alaska. Data was sourced from the Knowledge Network for Biocomplexity. Regional data: https://doi.org/10.5063/F1125QWP. Population data: https://doi.org/10.5063/F11G0JHX. River data: https://doi.org/10.5063/F1PZ573F.

### 2: Setup

load packages and data

```{r}

library(R.utils)
library(readr)
library(here)
library(sf)
library(ggplot2)
library(leaflet)
library(scales)
library(ggspatial)
library(dplyr)
library(prettymapr)

knb_url <- "https://dev.nceas.ucsb.edu/knb/d1/mn/v2/object/urn%3Auuid%3Aaceaecb2-1ce0-4d41-a839-d3607d32bb58"

#use libcurl dowloader to prevent corruption issues
download.file(knb_url,
              destfile = "shapefile_demo_data.zip",
              mode = "wb",
              method = "libcurl")

#unzip using the internal decompression instead of an external tool
unzip("shapefile_demo_data.zip", exdir = "data", unzip = "internal")

file.remove('shapefile_demo_data.zip')
```

### 3.0: Exploring the data

load data and set new CRS

```{r}
#read in shape file
ak_rgns_sf <- read_sf(here("data/ak_regions_simp.shp"))

#CRS set to Alaska Albers projection: 3338
ak_rgns_3338_sf <- ak_rgns_sf %>%
    st_transform(crs = 3338)

# read in population data
pop_df <- read_csv(here("data/alaska_population.csv"))
```

### 4.0: sf and the Tidyverse

Select for regions and filter by regions

```{r}
#select for region to add a region column to the spatial object
ak_rgns_3338_sf %>%
  select(region)

ak_rgns_3338_sf %>%
  filter(region == "Southeast")
```

### 5.0: Spatial Joins

Turn population data into a spatial object and join with geographic data

```{r}
#turn population data into a spatial object
pop_4326_sf <- st_as_sf(pop_df,
                        coords = c('lng', 'lat'),
                        crs = 4326,
                        remove = F)

#set the crs number of the population data object to 3338
pop_3338_sf <- st_transform(pop_4326_sf, 
                            crs = 3338)

#join population data with Alaska regions
pop_joined_sf <- st_join(pop_3338_sf,
                         ak_rgns_3338_sf,
                         join = st_within)

#find the total population by region
pop_rgn_df <- pop_joined_sf %>% 
  st_drop_geometry() %>% 
  group_by(region) %>% 
  summarize(total_pop = sum(population))

pop_rgn_3338_sf <- left_join(ak_rgns_3338_sf,
                             pop_rgn_df,
                             by ='region')

#save the spatial object with population data
write_sf(pop_rgn_3338_sf, here("data/ak_regions_population.shp"))

#plot pop_rgn_3338 to ensure that the joins worked
plot(pop_rgn_3338_sf["total_pop"])
```

### 6.0: Visualizing with ggplot

plotting with geom

```{r}
ggplot(pop_rgn_3338_sf) +
    geom_sf(aes(fill = total_pop)) +
    labs(fill = "Total Population") +
    scale_fill_continuous(low = "khaki",
                          high =  "firebrick",
                          labels = comma) +
    theme_bw()
```

Plot river locations

```{r}
#load in river data file
rivers_3338_sf <- read_sf(here("data/ak_rivers_simp.shp"))

ggplot() +
    geom_sf(data = pop_rgn_3338_sf, 
            aes(fill = total_pop)) +
    geom_sf(data = pop_3338_sf, 
            size = 0.5) +
    geom_sf(data = rivers_3338_sf,
            aes(linewidth = StrOrder)) +
    scale_linewidth(range = c(0.05, 0.5),
                    guide = "none") +
    labs(title = "Total Population by Alaska Region",
         fill = "Total Population") +
    scale_fill_continuous(low = "khaki",
                          high =  "firebrick",
                          labels = comma) +
    theme_bw()
```

Use ggspatial to incorporate base maps

```{r}
#set crs to 3587, the OpenStreetMap projection
pop_3857_sf <- st_transform(pop_3338_sf, 
                         crs = 3857)

#plot using ggspatial to add the openstreetmap tile layer
ggplot(data = pop_3857_sf) +
  ggspatial::annotation_map_tile(type = "osm", zoom = 4, progress = 'none') + 
  # higher zoom values are more detailed 
  geom_sf(aes(color = population),
          fill = NA) +
  scale_color_continuous(low = "darkkhaki",
                          high =  "firebrick",
                          labels = comma)
```

```{r}

```

### 7.0: Visualize with leaflet

```{r}
#create leaflet projection
epsg3338 <- leaflet::leafletCRS(
    crsClass = "L.Proj.CRS",
    code = "EPSG:3338",
    proj4def =  "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs",
    resolutions = 2 ^ (16:7)
)

#transform sf object crs back to 4326
pop_rgn_4326_sf <- pop_rgn_3338_sf %>% 
    st_transform(crs = 4326)

#create leaflet object with
pal <- colorNumeric(palette = "Reds", domain = pop_rgn_4326_sf$total_pop)

m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
    addPolygons(
        data = pop_rgn_4326_sf,
        fillColor = ~ pal(total_pop),
        weight = 1,
        color = "black",
        fillOpacity = 1
    ) %>%
    addCircleMarkers(
        data = pop_4326_sf,
        lat = ~ lat,
        lng = ~ lng,
        radius = ~ log(population / 500),
        # arbitrary scaling
        fillColor = "gray",
        fillOpacity = 1,
        weight = 0.25,
        color = "black",
        label = ~ paste0(pop_4326_sf$city, ", population ", comma(pop_4326_sf$population))
    ) %>%
    addLegend(
        position = "bottomleft",
        pal = pal,
        values = range(pop_rgn_4326_sf$total_pop),
        title = "Total Population"
    )

#show object
m
```

### Conclusions

Through visualization of Alaskan geographical regions and population density, it is plainly obvious that the vast majority of the state's population density is localized to the regions surrounding Anchorage, with some regions of higher population density in the areas surrounding Juneau and Fairbanks. The high population density of Anchorage makes sense given the sheltered waterfront provided by the inlet it sits on. However, the majority of Alaska has low population density beyond these few regions.
